Paradigms in Physics 2.0

### 2017-06-29 paper.tex
further refinement

\begin{abstract}
In 2016, the Department of Physics at Oregon State University began a
process to revise our Paradigms in Physics curriculum for physics
majors.  This paper will describe both the process by which our
department achieved consensus on this curricular change, and the
resulting curriculum.  The Paradigms 2.0 committee began with a
survey of students and faculty, followed by individual interviews with
the faculty teaching each existing course.  As we developed a plan to
address student- and faculty-identified challenges in the curriculum,
we met with each faculty member individually to explain and refine our
proposal, which was unanimously approved by the faculty.  Major
changes include major changes to several courses (math
methods, computational physics, modern physics, electronics, and
classical mechanics), including the introduction of two sophomore-year
courses designed specifically to help prepare students for their
upper-division courses.
\end{abstract}

\maketitle
% * <ethan.minot@oregonstate.edu> 2017-06-28T19:21:10.109Z:
%
% ^.
\section{Introduction}
\davidsays{Why does a reader care? A dramatic example of curriculum change.
A dramatic revision of said.  So others can adopt this method of change.  This is
an implementation of education research.  It is an exemplar for the community.
}
The Paradigms in Physics project began in 1996, when three faculty
members at Oregon State applied to the NSF for funding to redesign the
upper-division curriculum.  The motivation for the original Paradigms~1.0
project was in fact similar to the motivation for Paradigms~2.0: a
desire to soften the ``brick wall'' encountered by students upon
reaching the junior year, within the constraint that transfer students
must be able to graduate with two years of upper-division
courses~\cite{manogue2001paradigms}.  On top of this, there was a
desire to provide students with a broad knowledge of physics prior to
the GRE exams in the fall of their senior year.  The process of
redesigning the curriculum was dominated by the creation of index cards
listing subject content, and sorting these cards into courses.  This
process involved the entire faculty, and culminated in a unanimous
agreement to adopt the resulting curriculum.

@@

\subsection{Motivation for Paradigms 2.0}
There were a number of factors that motivated us to embark on the
Paradigms~2.0 process.  In the last two decades, we have observed
a number of challenges students face in our major.  While in many cases we
addressed these by changing and reordering the courses, we felt a look
at the entire curriculum was in order.  \davidsays{Triangulation of different sources.  Where did we get info about this issue?}
In addition, our faculty
understanding of the details of the sequence was diminishing by attrition: 
while our younger
faculty are enthusiastic about the Paradigms program, many lack perspective on
where students are in a given course, and what content is essential
for students to grasp for a subsequent course.

\paragraph{Transfer students}
Our curriculum is strongly affected by our desire to accommodate
transfer students from community colleges, who make up about a \davidsays{quarter}
of our Physics majors.  This leads us to focus on
a physics major in which all upper-division courses are taken in two
years.  However, our current curriculum is very hard on students in
the fall of their junior year, and especially so for transfer
students.  So we aimed to structure and order the courses such that non-transfer
students could be better prepared for their junior year, while
transfer students could postpone a few courses so that the
Fall quarter of their junior year would be no more difficult than that of any other student.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{vertical-schedule}
\caption{Hypothetical student schedule in the new curriculum.
  Transfer students begin in year 3, and take the marked additional
  courses.\label{fig:schedule}}
\end{figure}
\section{Changes made}
As a result of this process, we have implemented a number of changes
to our physics major curriculum.  The resulting curriculum is outlined
in Fig.~\ref{fig:schedule}.  The changes consisted of introducing two
new sophomore-level courses (which may be taken in the junior year) to
better prepare our majors for the their junior year.  We eliminated \emph{Modern
Physics} and the \emph{Classical Mechanics Capstone} in favor of the two new
sophomore-level courses.  We eliminated the \emph{Math Methods} course in
favor of ``\emph{Math Bits}'' integrated into the \emph{Paradigm} courses.
Finally, we restructured our nine 3-week \emph{Paradigm} courses into six
5-week \emph{Paradigm} courses.  We changed our computational physics requirement,
and reduced the number of electronics courses.

\subsection{5-week \emph{Paradigm} courses}
\davidsays{Advising flexibility, student focus, cohort building---daily,
active engagement.}
To begin with the most distinctive feature of our curriculum, we
wanted to maintain the existing intensive 7-hour-per-week schedule in
the junior year, which we have found effective.  However, we chose to
change from three 3-week courses per quarter to two 5-week courses
per quarter.    This change has several advantages.  \davidsays{Allows opportunity for
formative assessment and feedback for students.  Gives a chance for students to
catch up who fall behind.  Time to adjust to a new professor.  Simplifies teaching
loads} It gives students
a bit more time with each professor prior to their final exam.  This
gives more flexibility, for instance, in case of illness.
Finally, by reducing the number of courses, we allow the curriculum
to be taught with fewer faculty, albeit at an increased work load per professor.

>>>>

\begin{abstract}
In 2016, the Department of Physics at Oregon State University began a
process to revise our Paradigms in Physics curriculum for physics
majors.  We began with a
survey of students and faculty, followed by individual interviews with
the faculty teaching each existing course.  As we developed a plan to
address student- and faculty-identified challenges in the curriculum,
we met with each faculty member individually to explain and refine our
proposal, which was unanimously approved by the faculty.  Major
changes include major changes to several courses (math
methods, computational physics, modern physics, electronics, and
classical mechanics), including the introduction of two sophomore-year
courses designed specifically to help prepare students for their
upper-division courses.
\end{abstract}

\maketitle
% * <ethan.minot@oregonstate.edu> 2017-06-28T19:21:10.109Z:
%
% ^.
%\section{Introduction}
%\davidsays{Why does a reader care? A dramatic example of curriculum change.
%A dramatic revision of said.  So others can adopt this method of change.  This is
%an implementation of education research.  It is an exemplar for the community.
%}
The Paradigms in Physics project began in 1996, when three faculty
members at Oregon State applied to the NSF for funding to redesign the
upper-division curriculum.  The motivation for the original Paradigms~1.0
project was in fact similar to the motivation for Paradigms~2.0: a
desire to soften the ``brick wall'' encountered by students upon
reaching the junior year, within the constraint that transfer students
must be able to graduate with two years of upper-division
courses~\cite{manogue2001paradigms}.  On top of this, there was a
desire to provide students with a broad knowledge of physics prior to
the GRE exams in the fall of their senior year.  The process of
redesigning the curriculum was dominated by the creation of index cards
listing subject content, and sorting these cards into courses.  This
process involved the entire faculty, and culminated in a unanimous
agreement to adopt the resulting curriculum.

@@

%\subsection{Motivation for Paradigms 2.0}
A number of factors motivated us to embark on the
Paradigms~2.0 process.  In the last two decades, we have observed
a number of challenges students face in our major.  While in many cases we
addressed these by changing and reordering the courses, we felt a look
at the entire curriculum was in order.
In addition, our faculty
understanding of the details of the sequence was diminishing by attrition: 
while our younger
faculty are enthusiastic about the Paradigms program, many lack perspective on
where students are in a given course, and what content is essential
for students to grasp for a subsequent course.

\paragraph{Transfer students}
Our curriculum is strongly affected by our desire to accommodate
transfer students from community colleges.  This leads us to focus on
a physics major in which all upper-division courses are taken in two
years.  However, our current curriculum is very hard on students in
the fall of their junior year, and especially so for transfer
students.  So we aimed to structure and order the courses such that non-transfer
students could be better prepared for their junior year, while
transfer students could postpone a few courses so that the
Fall quarter of their junior year would be no more difficult than that of any other student.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{vertical-schedule}
\caption{Hypothetical student schedule in the new curriculum.
  Transfer students begin in year 3, and take the marked additional
  courses.\label{fig:schedule}}
\end{figure}
\section{Changes made}
As a result of this process, we have implemented a number of changes
to our physics major curriculum.  The resulting curriculum is outlined
in Fig.~\ref{fig:schedule}.  The changes consisted of introducing two
new sophomore-level courses (which may be taken in the junior year) to
better prepare our majors for the their junior year.  We eliminated \emph{Modern
Physics} and the \emph{Classical Mechanics Capstone} in favor of the two new
sophomore-level courses.  We eliminated the \emph{Math Methods} course in
favor of \emph{Math Bits} integrated into the \emph{Paradigm} courses.
Finally, we restructured our nine 3-week \emph{Paradigm} courses into six
5-week \emph{Paradigm} courses.  We changed our computational physics requirement,
and reduced the number of electronics courses.

\subsection{5-week \emph{Paradigm} courses}
To begin with the most distinctive feature of our curriculum, we
wanted to maintain the existing intensive 7-hour-per-week schedule in
the junior year, which we have found effective.  This schedule allows students to focus
intensely on a single topic, it fosters the building of a cohort of students, and 
the daily schedule is helpful for active engagement.  We chose to
change from three 3-week courses per quarter to two 5-week courses
per quarter.  This gives students who fall behind a chance to catch up, and enables
us to give more feedback to students prior to the final exam.  Having 3-credit courses
simplifies the setting of teaching loads, and by reducing the number of courses, 
we allow the curriculum to be taught with fewer faculty, albeit at an increased work
